Standardized Pupil Analysis

### 2023-12-01 ASHA 2023 Analysis/standardized_pupil_analysis.Rmd
Started normalization-WIP

---
title: "ASHA 2023 Listening Effort Data Analysis"
author: "Micah E. Hirsch"
date: "2023-10-30"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

@@

The purpose of this document is to run preliminary analyses and visualizations for this project. These results were prepared specifically for a poster presentation at the the 2023 Annual American Speech Language Hearing Association Convention being held in Boston, MA in November 2023.

Any questions or comments regarding this project or analysis may be directed to Micah Hirsch, mhirsch@fsu.edu.

This analysis was conducted using R version 4.3.1

@@

# Descriptive Statistics

## Participant Demographics

First, we will present the demographics of our listeners To date, we recruited 20 listeners to participate in the study. Three listeners were removed from the analysis. One listener decided to withdraw from the study, so their data was removed from the database. For the other two participants, pupillometry data was either not recorded or lost due to instrumental issues. Therefore, for this analysis, we have 17 participants. The descriptive information for these participants are presented below.

```{r, warning = F, message = F}

# Load in dataset for listener demographics

listener_demo <- rio::import("Cleaned Data/cleaned_listener_demo.csv")

# Extracting listener gender data

gender <- listener_demo %>%
  dplyr::select(gender) %>%
  dplyr::group_by(gender) %>%
  dplyr::summarize(Frequency = n()) %>%
  dplyr::rename(Category = "gender") %>%
  dplyr::mutate(Group = "Gender",
                Category = fct_relevel(Category, c("woman", "man", "nonbinary"))) %>%
  dplyr::arrange(Category)

## Extracting listener race data

race <- listener_demo %>%
  dplyr::select(race) %>%
  dplyr::group_by(race) %>%
  dplyr::summarize(Frequency = n()) %>%
  dplyr::rename(Category = "race") %>%
  dplyr::mutate(Group = "Race",
                Category = fct_relevel(Category, c("white/Caucasian", "Biracial or Multiracial", 
                                                   "race not listed"))) %>%
  dplyr::arrange(Category)

## Extracting listener ethnicity data

ethnicity <- listener_demo %>%
  dplyr::select(ethnicity) %>%
  dplyr::group_by(ethnicity) %>%
  dplyr::summarize(Frequency = n()) %>%
  dplyr::rename(Category = "ethnicity") %>%
  dplyr::mutate(Group = "Ethnicity",
                Category = fct_relevel(Category, c("Hispanic/Latino(a/e)", 
                                                   "Not Hispanic/Latino(a/e)"))) %>%
  dplyr::arrange(Category)

# Creating Demographic Table 1

dplyr::full_join(gender, race) %>%
  dplyr::full_join(., ethnicity) %>%
  # Initiating a gt table
  gt::gt(rowname_col = "Category",
         groupname_col = "Group")

@@

As indicated in the table above, our participants were primarily white/Caucasian, women, and not Hispanic or Latino(a/e). All participants reported no history of a speech, language, or hearing disorder. All participants were fluent English speakers. However, two indicated their native language was different than American English. These languages were Spanish (n=1) and Turkish (n=1).

```{r, warning = F, message = F}

listener_demo %>%
  dplyr::select(c(age)) %>%
  dplyr::summarize(Mean = mean(age), SD = sd(age),
                   Min = min(age), Max = max(age)) %>%
  dplyr::mutate(Category = "Age") %>%
  dplyr::relocate(Category, .before = Mean) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(Mean, SD),
                decimals = 2) %>%
  gt::cols_label(Category = "")

The mean age of the listeners is 26.53 years old. Listener age ranged from 18 to 37 years old.

# Removing unneeded items from the environment

rm(ethnicity, gender, listener_demo, race)

## Speaker Intelligibility and Phrase Accuracy

### Mean Speaker Intelligibility

Across all trials in the study, the intelligibility level for each of the speaker is reported below.  Across listeners, the mean intelligibility for the ALS speaker was 81.08% with a standard deviation of 3.86. For the control speaker, the mean intelligibility level was 93.62% with a standard deviation of 2.90. Based on their intelligibility level, the speaker with ALS is determined to have a moderate severity of dysarthria (Stipancic et al., 2021).

```{r, warning = F, message = F}

# Load in cleaned data for pupil and perceived listening effort data

data <- rio::import("Cleaned Data/cleaned_data.csv") %>%
  # removing trials with missing rep accuracy: 3 trials removed
  dplyr::mutate(rep_acc = ifelse(rep_acc == "", NA, rep_acc)) %>%
  dplyr::filter(!is.na(rep_acc)) %>%
  # Releveling speaker variable
  dplyr::mutate(speaker = as.factor(speaker),
                speaker = fct_relevel(speaker, "Control", "ALS")) %>%
  # Creating a trial variable that codes trial 6 (i.e. the first experimental trial) as 0.
  dplyr::mutate(trial_c = trial - 6)

# Selecting variables needed to calculate mean speaker intelligibility
intel_df <- data %>%
  dplyr::select(subject, trial, speaker, correct_words, target_number, rep_acc) %>%
  dplyr::distinct()

# Calculating intelligibility for the control and ALS speaker for each listener
intel_byListener <- intel_df %>%
  dplyr::group_by(speaker, subject) %>%
  dplyr::summarize(Intelligibility = sum(correct_words)/sum(target_number)*100)

intel_byListener %>%
  dplyr::group_by(speaker) %>%
  dplyr::summarize("Mean Intelligibility" = mean(Intelligibility), SD = sd(Intelligibility)) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c("Mean Intelligibility", SD),
                 decimals = 2) %>%
  gt::cols_label(speaker = "Speaker")

### Phrase Accuracy

For this analysis, we are only looking at the trials that listeners were able to accurately recognize (e.g. 100% intelligible to the listener). Therefore, we are looking at the proportion of trials for each speaker that listeners were able to accurately recognize compared to the total number of trials. Although our ALS speaker's intelligibility was fairly high, only about 49% of all the ALS speaker's trials were accurately recognized across listeners.

```{r, warning = F, message = F}

# Calculating repetition accuracy descriptive data

intel_df %>%
  dplyr::group_by(speaker, rep_acc) %>%
  dplyr::summarize(n = n()) %>%
  tidyr::pivot_wider(names_from = rep_acc,
                     values_from = n) %>%
  dplyr::mutate(percent_accurate = accurate/(accurate + inaccurate)*100) %>%
  dplyr::ungroup() %>%
  gt::gt(rowname_col = "speaker") %>%
  gt::fmt_number(columns = percent_accurate,
                 decimals = 2) %>%
  gt::cols_label(accurate = "Accurate",
                 inaccurate = "Inaccurate",
                 percent_accurate = "Percent Accurate")

# removing unneeded items from the environment

rm(intel_byListener, intel_df)

# Perceived Listening Effort Ratings

## Descriptives

```{r, warning = F, message = F}

perceived_effort_df <- data %>%
  dplyr::select(subject, trial, speaker, effort_rating, rep_acc, code, trial_c) %>%
  dplyr::distinct() %>%
  # For this particular analysis, we are just looking at results for accurate repetition
  # However, the final project will look at rep_acc too
  dplyr::filter(rep_acc == "accurate")

per_effort_des <- perceived_effort_df %>%
  dplyr::group_by(speaker) %>%
  dplyr::summarize(per_effort = mean(effort_rating), 
                   sd = sd(effort_rating, na.rm = T),
                   se = sd/sqrt(n()))

per_effort_des %>%
  gt::gt() %>%
  gt::fmt_number(columns = per_effort:se,
                 decimals = 2) %>%
  gt::cols_label(speaker = "Speaker",
                  per_effort = "Perceived Effort",
                  sd = "SD",
                  se = "SE")

As reported in the table above and shown in the plot below, the mean perceived effort ratings for the ALS speaker is about 2.13 points higher than the control speaker. This is reflecting the difference in mean perceived effort ratings between the ALS and control speaker, even when the phrase was intelligible to the listener.

```{r, warning = F, message = F}

# Creating a color palette vector for plots
my_pal <- c("#CD9FCC", "#7E2E84")

# Plotting the mean perceived effort ratings for the ALS and control speaker
per_effort_des %>%
  ggplot() +
   aes(x = speaker,
      y = per_effort,
      group = speaker,
      color = speaker,
      fill = speaker) +
  geom_bar(stat = "identity", alpha = 0.6) +
  geom_errorbar(aes(x = speaker, ymin = per_effort - se, ymax = per_effort + se), width = 0.4) +
  labs(x = "Speaker", y = "Mean Perceived Listening Effort Rating") +
  theme_bw() +
  theme(legend.position = "none", aspect.ratio = 1) +
  scale_color_manual(values = my_pal) +
  scale_fill_manual(values = my_pal)

# Saving the ggplot

ggsave("Plots/Perceived_effort.png", plot = last_plot(), height = 5, width = 5, units = "in")

```{r, warning = F, message = F}

perceived_effort_df %>%
  ggplot() +
  aes(y = speaker,
      x = effort_rating,
      color = speaker,
      fill = speaker) +
  geom_density_ridges(scale = 0.7, alpha = 0.5) +
  labs(x = "Perceived Listening Effort Rating", y = "Speaker") +
  theme_bw() +
   theme(legend.position = "none", aspect.ratio = 1) +
  scale_fill_manual(values = my_pal) +
  scale_color_manual(values = my_pal)

# Saving the ggplot

ggsave("Plots/Perceived_effort_dis.png", plot = last_plot(), height = 5, width = 5, units = "in")

## Perceived Listening Effort Model

### Fully Unconditional Model

To analyze the perceived effort ratings, we are going to use linear mixed-effects models. To start, we are going to specify our fully unconditional model. In this model, we included random intercepts for subject (i.e. our listeners) and code (i.e. the different phrases). No fixed effects have been added to the model. We also calculated ICC from this model.

Both the conditional and unconditional ICC is .099 (rounded to .1 in model output).

```{r, warning = F, message = F}

m0 <- glmmTMB(effort_rating ~ 1 + (1|subject) + (1|code), data = perceived_effort_df)

sjPlot::tab_model(m0, pred.labels = "Intercept",
                  dv.labels = "Perceived Listening Effort")

performance::icc(m0)

### Model 1

Now we will add trial number (i.e. trial_c) as a random slope for both subject and code. A likelihood ratio test was also ran to test whether model 1 provided a better fit to the data than the unconditional model.

```{r, warning = F, message = F}

m1 <- glmmTMB(effort_rating ~ 1 + (trial_c|subject) + (trial_c|code), 
           data = perceived_effort_df)

sjPlot::tab_model(m1, pred.labels = "Intercept",
                  dv.labels = "Perceived Listening Effort")

anova(m0, m1)

The likelihood ratio test indicated that model 1 provided a better fit to the data than the unconditional model (chi-sq(4) = 227.79, p <0.001). Therefore, we will retain model 1.

### Model 2

To control for trial order, the trial order variable was added as a fixed effect to model 2. A likelihood ratio test was also conducted to test whether model 2 provides a better fit to the data compared to model 1

```{r, warning = F, message = F}

m2 <- glmmTMB(effort_rating ~ trial_c + (trial_c|subject) + (trial_c|code), 
              data = perceived_effort_df)

sjPlot::tab_model(m2, pred.labels = c("Intercept", "Trial Order"),
                  dv.labels = "Perceived Listening Effort")

anova(m1, m2)

As indicated from the model, adding trial order did not significantly improve the model fit to the data (chi-sq(1) = .412, p = .52). This indicates that trial order is not an important predictor for perceived listening effort ratings and that ratings for the speakers were fairly consistent across trials.

### Model 3

In model 3, we added the speaker fixed effect to test whether the mean difference in perceived listening effort rating is significantly difference for the ALS speaker compared to the control speaker. A likelihood ratio test was also conducted to test whether model 3 provides a better fit to the data compared to model 2.

```{r, warning = F, message = F}

m3 <- glmmTMB(effort_rating ~ trial_c + speaker + (trial_c|subject) + (trial_c|code), 
           data = perceived_effort_df)

sjPlot::tab_model(m3, pred.labels = c("Intercept", "Trial Order", "Speaker [ALS]"),
                  dv.labels = "Perceived Listening Effort")

anova(m2, m3)

Based on the model output, adding speaker to the model improved the model fit to the data (chi-sq(1) = 469.74, p <.001). The fixed effect estimate indicates that, while controlling for trial order, the perceived listening effort rating across listeners and phrases is about 2.35 higher for the ALS speaker.

### Model 4

To test whether there are any potential differences in perceived effort ratings across trials, we are testing whether adding an interaction between trial order and speaker improves model fit. Again, a likelihood ratio test was conducted to see if model 4 provides a better fit to the data compared to model 3.

```{r, warning = F, message = F}

m4 <- glmmTMB(effort_rating ~ trial_c + speaker + trial_c*speaker + (trial_c|subject) + (trial_c|code), 
           data = perceived_effort_df)

sjPlot::tab_model(m4, pred.labels = c("Intercept", "Trial Order", "Speaker [ALS]", "Trial Order * Speaker [ALS]"),
                  dv.labels = "Perceived Listening Effort")

anova(m3, m4)

Based on the model output, adding the interaction term did not significantly improve model fit (chi-sq(1) = .14, p= .70). The fixed effect estimate for the interaction term is also not significant. This indicates that effort ratings across the trials for both the ALS speaker and the control speaker were not significantly different from each other. In other words, the ratings for both the ALS speaker and control speaker stayed consistent across the experimental trials.

### Final Model

Based on our model building, we will now specify our final model. Since the trial fixed effect as well as the interaction between trial and speaker were not significant, nor did they improve model fit, they were removed from the final model. All other output remained the same.

@@

per_effort_model_final <- glmmTMB(effort_rating ~ speaker + (trial_c|subject) + (trial_c|code), 
                               data = perceived_effort_df)

sjPlot::tab_model(per_effort_model_final, pred.labels = c("Intercept", "Speaker [ALS]"),
                  dv.labels = "Perceived Listening Effort",
                  file = "Final Model Outputs/perceived_effort_model.html")
```

The final model indicates that the perceived effort ratings for the ALS speaker is about 2.35 points higher than the control speaker. This is consistent with the results shown in the descriptive statistics.

# Remove unneeded items from the environment

rm(m0, m1, m2, m3, m4, per_effort_des, perceived_effort_df, per_effort_model_final)

# Pupil Dilation

## Descriptives

The ALS speaker has a slower speech rate compared to the control speaker. Therefore, we are going to calculate the average phrase length to see on average, how different the phrase lengths were for ALS trials and control speaker trials. To calculate this, the first 3000 ms were filtered out of the df to remove the trial time before the phrase onset. Then, the average trial length was calculated (phrase presentation + 3000 ms retention period). The average phrase offset was calculated by subtracting 2000 ms from the average trial length, since our analysis region of interests stops 1000 ms after the phrase offset.

We will need this information to be able to mark the average phrase offset time and average end point for the analysis region of interest for visualization purposes.

```{r, warning = F, message = F}

@@

On average, the phrase length (phrase presentation), was about 2461 ms longer for the ALS trials than the control speaker trials.

### Visualization

#### Average Pupil Response to the Control Speaker

This is the smoothed pupil dilation tracking for the Control Speaker. The solid line indicates the start of the phrase. The dotted line indicates the average phrase offset. The grey box represents the region of interest for analysis.

```{r, warning = F, message = F}

control <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "Control") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1908, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2908, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,4908), ylim = c(-100, 200)) +
  labs(title = "Control Speaker", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[1]) +
  theme(legend.position = "none", aspect.ratio = 1)

control

#### Individual Listener Pupil Response to the Control Speaker

We can also look at this tracking grouped by the listeners in our study. From this figure, we can see there are variable responses to the the control speaker across listeners.

```{r, warning = F, message = F}

control_byListener <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "Control") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      group = subject,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1908, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2908, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,4910), ylim = c(-100, 200)) +
  labs(title = "Pupil Response to Control Speaker (Grouped by Listener)", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[1]) +
  theme(legend.position = "none", aspect.ratio = 1)

control_byListener

#### Individual Listener Pupil Response to the ALS Speaker

This is the smoothed pupil dilation tracking for the ALS Speaker. The solid line indicates the start of the phrase. The dotted line indicates the average phrase offset. The grey box represents the region of interest for analysis.

```{r, warning = F, message = F}

als <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "ALS") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 4369, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 5369, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,7369), ylim = c(-100, 200)) +
  labs(title = "ALS Speaker", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[2]) +
  theme(legend.position = "none", aspect.ratio = 1)

als

#### Average Pupil Response to the ALS Speaker

We can also look at this tracking grouped by the listeners in our study. From this figure, we can see there are variable responses to the the  speaker with ALS across listeners. Some of listener trackings for this speaker are following an odd pattern. Future analyses will investigate these individual listener and trial trackings to check for any potential tracking errors.

```{r, warning = F, message = F}

als_byListener <- data %>%
  dplyr::filter(rep_acc == "accurate") %>%
  dplyr::filter(speaker == "ALS") %>%
  ggplot() +
  aes(x = timebins,
      y = pupil.binned,
      group = subject,
      color = speaker) +
  geom_smooth() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 4369, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 5369, ymin = -50, ymax = 150, alpha = 0.2) +
  coord_cartesian(xlim = c(0,7369), ylim = c(-100, 200)) +
  labs(title = "ALS Speaker", x = "Time (ms)", y = "Pupil Dilation (arbitrary units)") +
  theme_bw() +
  scale_color_manual(values = my_pal[2]) +
  theme(legend.position = "none", aspect.ratio = 1)

als_byListener

pupil_plot <- ggarrange(control, als, ncol = 2)

ggsave("Plots/Pupil_dilation.png", plot = pupil_plot, width = 10, height = 5, units = "in")

rm(als, control, als_byListener, control_byListener)

## Pupil Dilation Analysis

### Unconditional Growth Model

First, we ran an unconditional growth model. In the intial model, I first specified a model with the orthogonal polynomials added as fixed effects and subject (i.e. listener) and code (i.e. phrase) as random intercepts. I also added the orthogonal polynomials as random slopes (poly1, poly2, and poly3). I also attempted to use the AR(1) covariance structure to specifiy the random effect structure. However, I was not able to run the model and ran into convergence issues.

Therefore, I decided to respecify the random effect structure by removing the AR(1) specification. This led to a model with singular fit. In response to this, I decided to remove code (i.e.phrase) from the random effect structure in order to simplify it.

For more details on how the unconditional growth model was specified, please click on "show" to see the code.

The adjusted ICC for the unconditional growth model is .109 (rounded to .11 in model output). The unadjusted ICC for the conditional growth model is .107.

```{r, warning = F, message = F}

# Specifying the region of interest for analysis (500 ms to 1000 ms after relative offset of the trial)

pupil_roi <- data %>%
  # For this analysis, we are only analyzing pupil dilation for accurate responses
  dplyr::filter(rep_acc == "accurate") %>%
  # filtering out timebins below 500 (i.e., keeping timebins above 500 ms)
  dplyr::filter(timebins >= 500) %>%
  dplyr::group_by(subject, trial) %>%
  dplyr::filter(timebins <= max(timebins) - 2000) %>%
  ungroup()

# Adding linear, quadratic, and cubic orthogonal polynomials to the df

pupil_roi <- gazer::code_poly(pupil_roi, predictor = "timebins", poly.order = 3, orthogonal = T, draw.poly = F)

# Exporting region of interest df (this can be used to )

rio::export(pupil_roi, "Cleaned Data/pupil_analysis_roi.csv")

# I kept running into issues using glmmTMB. Therefore, I switched to using lme4 for these analyses

# specifying unconditional growth model for pupil data

m0 <- lmer(pupil.binned ~ poly1 + poly2 + poly3 +
                (poly1 + poly2 + poly3|subject), 
              data = pupil_roi)

sjPlot::tab_model(m0, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term"),
                  dv.labels = "Pupil Dilation")

performance::icc(m0)

### Model 1

In the next model, we are adding a fixed effect for speaker to test whether on average, pupil dilation is greater for the ALS speaker compared to the control speaker. A likelihood ratio test was also conducted to test whether model 1 provides a better fit to the data than the unconditional growth model.

```{r, warning = F, message = F}

m1 <- lmer(pupil.binned ~ poly1 + poly2 + poly3 + speaker +
                (poly1 + poly2 + poly3|subject), 
              data = pupil_roi)

sjPlot::tab_model(m1, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term", "Speaker [ALS]"),
                  dv.labels = "Pupil Dilation")

anova(m0, m1)

As indicated in the model output, when listening to the ALS speaker, on average, listeners had greater pupil dilation compared to when they were listening to the control speaker. The likelihood ratio test also indicates model 1 is a better fit to the data compared to the unconditional model (chi-sq(1) = 353.67, p <0.001).

### Model 2

In model two, the interaction between the linear growth term and speaker was added to the model as a fixed effect. A likelihood ratio test was also conducted to test if model 2 provides a better fit to the data compared to model 1.

Note: The initial run of this model did not converge, so an optimizer control was added to the model. This optimizer will remain in each subsequent model.

```{r, warning = F, message = F}

# Model 2 initially failed to converge, so adding a control statement for bobyqa optimizer

control_bobyqa <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e9))

m2 <- lmer(pupil.binned ~ poly1 + poly2 + poly3 + speaker +
             poly1*speaker +
             (poly1 + poly2 + poly3|subject), 
              data = pupil_roi, control = control_bobyqa)

sjPlot::tab_model(m2, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term", "Speaker [ALS]", 
                                      "Linear Term * Speaker [ALS]"),
                  dv.labels = "Pupil Dilation")

anova(m1, m2)

Based on the model output, the interaction between the linear growth term and speaker is significant, indicating that, across listeners, pupil dilation grew at a faster rate for the ALS speaker compared to the control speaker. The likelihood ratio test also indicated that model 2 provided a better fit to the data than model 1 (chi-sq(1) = 649.22, p <.001).

### Model 3

In model three, the interaction between the quadratic growth term and speaker was added to the model. This tests whether the quadratic shape of pupil dilation is significantly different between the two speakers. A likelihood ratio test was also conducted to test whether model 3 provides a better fit to the data compared to model 2.

```{r, warning = F, message = F}

m3 <- lmer(pupil.binned ~ poly1 + poly2 + poly3 + speaker +
             poly1*speaker + poly2*speaker +
             (poly1 + poly2 + poly3|subject), 
              data = pupil_roi, control = control_bobyqa)

sjPlot::tab_model(m3, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term", "Speaker [ALS]", 
                                      "Linear Term * Speaker [ALS]", "Quadratic Term * Speaker [ALS]"),
                  dv.labels = "Pupil Dilation")

anova(m2, m3)

Based on the model output, the interaction between the quadratic growth term and speaker improved the model fit (ch-sq(1) = 147.72, p<.001). This indicates, on average, the shape of the listeners' pupil dilation is significantly different while listening to the ALS speaker vs the control speaker.

### Model 4

In model for, we added the interaction between the cubic growth term and speaker. As with the other models, a likelihood ratio test was conducted to test whether model 4 provides a better fit to the data compared to model 3.

```{r, message = F, warning = F}

m4 <- lmer(pupil.binned ~ poly1 + poly2 + poly3 + speaker +
             poly1*speaker + poly2*speaker + poly3*speaker +
             (poly1 + poly2 + poly3|subject), data = pupil_roi, 
           control = control_bobyqa)

sjPlot::tab_model(m4, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term", "Speaker [ALS]", 
                                      "Linear Term * Speaker [ALS]", "Quadratic Term * Speaker [ALS]", "Cubic Term * Speaker [ALS]"),
                  dv.labels = "Pupil Dilation")

anova(m3, m4)

Based on the model output, model 4 provides a better fit to the data compared to model 3 (chi-sq(1) = 9.25, P = .002). Interestingly, the only fixed effects that are significant in this model are the main effect of the cubic growth term and the interaction between the cubic growth term and speaker. However, the model output still indicates that there are significant differences in the shape of pupil dilation for the ALS speaker compared to the control speaker.

```{r, include = F, message = F, warning = F}

# Saving model 4 output

sjPlot::tab_model(m1, m2, m3, m4, pred.labels = c("Intercept", "Linear Term", "Quadratic Term", "Cubic Term", "Speaker [ALS]", 
                                      "Linear Term * Speaker [ALS]", "Quadratic Term * Speaker [ALS]", "Cubic Term * Speaker [ALS]"),
                  dv.labels = c("Model 1", "Model 2", "Model 3", "Model 4"),
                  file = "Final Model Outputs/pupil_dilation_model.html")

# removing unneeded items from the environment

rm(m0, m1, m2, m3, m4, pupil_plot, control_bobyqa)

# Alternative Pupil Dilation Analysis

Since the trials for the ALS speaker are much longer than the control speaker's trials, we are going to conduct an exploratory analysis for the pupil dilation measure. For each of the listener's trial, we are going to extract the peak pupil dilation during the analysis region and model it as a function of speaker.

## Peak Pupil Dilation Descriptives

The mean peak pupil dilation, standard deviation, and standard error was calculated for each speaker and are reported below. In general, the mean peak pupil dilation for the ALS speaker is about 69 units higher than for the control speaker.

```{r, warning = F, message = F}

peak_pupil_df <- pupil_roi %>%
  dplyr::group_by(subject, trial_c, speaker, code) %>%
  dplyr::summarize(peak_pupil = max(pupil.binned)) %>%
  ungroup()

peak_pupil_des <- peak_pupil_df %>%
  dplyr::group_by(speaker) %>%
  dplyr::summarize(pupil = mean(peak_pupil), 
                   sd = sd(peak_pupil, na.rm = T),
                   se = sd/sqrt(n()))

peak_pupil_des %>%
  gt::gt() %>%
  gt::fmt_number(columns = pupil:se,
                 decimals = 2) %>%
  gt::cols_label(speaker = "Speaker",
                  pupil = "Peak Pupil Dilation",
                  sd = "SD",
                  se = "SE")

### Visualization

Below are two figures to visualize the distribution of peak pupil dilation across listeners and trials for the ALS and Control Speakers.

#### Peak Pupil Dilation Distribution Plot

```{r, warning = F, message = F}

peak_pupil_df %>%
  ggplot() +
  aes(y = speaker,
      x = peak_pupil,
      color = speaker,
      fill = speaker) +
  geom_density_ridges(scale = 0.7, alpha = 0.5) +
  labs(x = "Mean Peak Pupil Dilation (arbitrary units)", y = "Speaker") +
  theme_bw() +
  theme(legend.position = "none", aspect.ratio = 1) +
  scale_fill_manual(values = my_pal) +
  scale_color_manual(values = my_pal)

ggsave("Plots/peak_pupil_dis.png", plot = last_plot(), height = 5, width = 5, units = "in")

#### Peak Pupil Dilation Bar Plot

```{r, warning = F, message = F}

peak_pupil_des %>%
  ggplot() +
   aes(x = speaker,
      y = pupil,
      group = speaker,
      color = speaker,
      fill = speaker) +
  geom_bar(stat = "identity", alpha = 0.6) +
  geom_errorbar(aes(x = speaker, ymin = pupil - se, ymax = pupil + se), width = 0.4) +
  labs(x = "Speaker", y = "Mean Peak Pupil Dilation (arbitrary units)") +
  theme_bw() +
  theme(legend.position = "none", aspect.ratio = 1) +
  scale_color_manual(values = my_pal) +
  scale_fill_manual(values = my_pal)

ggsave("Plots/peak_pupil_bar.png", plot = last_plot(), height = 5, width = 5, units = "in")

## Analysis

### Fully Unconditional Model

We first fit a fully unconditional model to the data to calculate ICC. The adjusted and unadjusted ICC is 0.364.

```{r, warning = F, message = F}

# Note: I ran into issues using glmmTMB, so used lme4 instead

m0 <- lmer(peak_pupil ~ 1 + (1|subject) + (1|code), data = peak_pupil_df)

sjPlot::tab_model(m0, pred.labels = "Intercept",
                  dv.labels = "Peak Pupil Dilation")

performance::icc(m0)

### Model 1

We added trial as random slopes for both the subject and code random effects. The model initially failed to converge, so an optimizer was added to the model. Then a singular fit warning was displayed, so the random slope of trial_c was removed from the code random effect.

```{r, warning = F, message = F}

# Model failed to converge, so added an optimizer
# Singular fit so removed random slope for code.

m1 <- lmer(peak_pupil ~ 1 + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

sjPlot::tab_model(m1, pred.labels = "Intercept",
                  dv.labels = "Peak Pupil Dilation")

anova(m0, m1)

Model 1 provided a better fit to the data (chi-sq(2) = 43.44, p < .001).

### Model 2

Next, in order to control for trial order, trial was added to the model as a fixed effect.

```{r, warning = F, message = F}

m2 <- lmer(peak_pupil ~ trial_c + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

sjPlot::tab_model(m2, pred.labels = c("Intercept", "Trial Order"),
                  dv.labels = "Peak Pupil Dilation")

anova(m1, m2)

Adding trial to the model as a fixed effect significantly improved model fit (chi-sq(1) = 12.73, p<.001). Based on the fixed effect estimate for trial order, peak pupil dilation decreased approximately -1.23 units on average with each trial.

### Model 3

Next, we added the fixed effect of speaker to the model to test whether peak pupil dilation was, on average, higher for when listening to the ALS speaker compared to the control speaker.

```{r, warning = F, message = F}

m3 <- lmer(peak_pupil ~ trial_c + speaker + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

sjPlot::tab_model(m3, pred.labels = c("Intercept", "Trial Order", "Speaker [ALS]"),
                  dv.labels = "Peak Pupil Dilation")

anova(m2, m3)

Adding speaker to the model improved the model fit to the data (ch-sq(1) = 102.96, p <.001). Controlling for trial order, listeners' peak pupil dilation, on average, was about 85.22 units higher compared to when they were listening to the control speaker. This indicates listeners were exerting more effort while listening to the ALS speaker compared to the control speaker.

### Model 4

Finally, the interaction between trial and speaker was added to the model to test whether peak pupil dilation for the ALS speaker change at a different rate across trials.

```{r, warning = F, message = F}

# Ran into convergence problem, so removed the trial_c random slope for code

m4 <- lmer(peak_pupil ~ trial_c + speaker + speaker*trial_c + (trial_c|subject) + (1|code), 
              data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

sjPlot::tab_model(m4, pred.labels = c("Intercept", "Trial Order", "Speaker [ALS]", "Trial Order * Speaker [ALS]"),
                  dv.labels = "Peak Pupil Dilation")

anova(m3, m4)

Based on the model output, adding the speaker*trial interaction did not significantly improve model fit to the data (chi-sq(1) = 0.06, p = .80). This indicates that when listening to the speaker with ALS, the listeners' peak pupil dilation declined at a similar rate as when they were listening to the control speaker.

#### Final Model

Based on the model comparisons, model 3 provided the best fit to the data. Therefore, the final model output for peak pupil dilation is provided below.

```{r, warning = F, message = F}

final_model <- lmer(peak_pupil ~ trial_c + speaker + (trial_c|subject) + (1|code), 
              data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

sjPlot::tab_model(final_model, pred.labels = c("Intercept", "Trial Order", "Speaker [ALS]"),
                  dv.labels = "Peak Pupil Dilation",
                  file = "Final Model Outputs/peak_pupil_model.html")

# Removing unneeded items from the environment

rm(data, m0, m1, m2, m3, m4, final_model, peak_pupil_des, peak_pupil_df, pupil_roi, my_pal)

>>>>

---
title: "Standardized Pupil Analysis"
author: "Micah E. Hirsch"
date: "2023-12-01"
output: html_document
---

@@

The preliminary pupillometry results (found in the data_analysis.Rmd file) indicicated that the shape of the pupil dilation tracking is different while listening to the speaker with ALS verses listening to the control speaker. However, it is not clear whether this finidng is influenced by the difference in the length of the audio files for the control speaker and the ALS speaker. Due to the speaker's slowed speech rate, the ALS speaker's audio files were longer than the control speaker's files.

Therefore, in order to control for the effect of the difference in speech rate, we will normalize the trial-level data for the ALS speaker so that the trial lengths are similar to the control speaker's trials.

This reanalysis is conducted using R version 4.3.1

@@

# Pupil Dilation Normalization

Work in Progress

# Import pupil dilation data
data <- rio::import("Cleaned Data/cleaned_data.csv") %>%
  # removing trials with missing rep accuracy: 3 trials removed
  dplyr::mutate(rep_acc = ifelse(rep_acc == "", NA, rep_acc)) %>%
  dplyr::filter(!is.na(rep_acc)) %>%
  # Releveling speaker variable
  dplyr::mutate(speaker = as.factor(speaker),
                speaker = fct_relevel(speaker, "Control", "ALS")) %>%
  # Creating a trial variable that codes trial 6 (i.e. the first experimental trial) as 0.
  dplyr::mutate(trial_c = trial - 6)

# Pupil Normalization

bin.length = 2.29

ALS_trials <- data %>%
  dplyr::filter(speaker == "ALS") %>%
  dplyr::mutate(time_n = round(timebins/bin.length)*bin.length) %>%
  dplyr::group_by(subject, trial_c, speaker, time_n, code) %>%
  dplyr::summarize(normed_pupil = mean(pupil.binned)) %>%
  dplyr::ungroup()

@@

@@

@@
